---
title: "CosMx Cell QC"
author: Amélie Julé (Sansom Lab)
date: "`r Sys.Date()`"
params:
  group: NULL
  cell_nCount_cutoff: NULL
  cell_nFeature_cutoff: NULL
  cell_percentNeg_cutoff: NULL
  cell_complexity_cutoff: NULL
  exclCellsInFailedFOVs: NULL
  sampleMinCells: NULL
  sampleMinPercent: NULL
  composite_ncol: 2
  composite_spacing: 500
  plot_ncol_spatial: 4
  plot_ncol: 2
  font_size: 10
  legend_size: 3
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include = FALSE} 
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = TRUE,
                      prompt = FALSE, cache = FALSE, tidy = FALSE)
set.seed(123456789)
report <- "cellQC"
```

```{r libraries}
library(data.table)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(patchwork)
library(RColorBrewer)
```

```{r echo=FALSE, eval=FALSE, include = FALSE}
# For debugging of Snakemake workflow only
print(getwd())
print(params)
```

```{r graphical_params, child="_graphical_params.Rmd", eval=file.exists("_graphical_params.Rmd")}
```

```{r header, child="probeqc_header.Rmd", eval=file.exists("probeqc_header.Rmd")}
```


## Slide overview

**The following sample(s) are found on each slide:**

```{r, fig.height=12L}
# Generate list of spatial plots per slide
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = sample_id)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_manual(values = sample(gg_color_hue(length(unique(spx_meta$sample_id))))) +
    guides(colour = guide_legend(override.aes = list(size = legend_size))) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed() #+ theme(legend.position = 'none')
    #geom_hline(yintercept = unique(fov_df$y_global_px[fov_df$FOV %in% fovs]), col = "grey50") +
    #geom_vline(xintercept = unique(fov_df$x_global_px[fov_df$FOV %in% fovs]), col = "grey50")
  
  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol_spatial))
```


Before QC, each sample totals the following number of cells:

```{r}
qc_sample <- spx_meta[, c("sample_id", "nCell_sampleSum")]
qc_sample <- qc_sample[!duplicated(qc_sample), ]
rownames(qc_sample) <- 1:nrow(qc_sample)
qc_sample |> dplyr::arrange(desc(nCell_sampleSum))
```

* * *
* * *
* * *


# Cell QC

"*Cell QC flags cells with low or spurious signal based on the number of targets detected (more is better), fraction of probes which are negative controls (fewer is better), uniformity of count distribution (higher complexity is preferred), and cell size (the top percentile may need to be removed as a QC of segmentation).*" (from AtoMx user manual) 

**NOTE: `spatialhub` leaves the cell size QC parameter aside -- this is considered as a segmentation QC.**


## Visual assessment of variation in space

The following plots highlight the areas with higher `nCount_RNA` per cell, which tends to be sample/morphology specific. **Color scales are capped for contrast, readability and comparability across slides (where applicable).**


### Total counts per cell (all probes)

```{r, fig.height=12L}
col_max <- signif(max(spx_meta$nCount_RNA) * 2/3, 2)
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(nCount_RNA) |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = nCount_RNA)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_gradient(low = "grey75", high = "blue",
                         limits = c(0, col_max), oob = scales::squish) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed() + theme(legend.position = 'none')
  
  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol_spatial))
```


### Total counts and distinct probes per cell: per-sample Zoom {.tabset}

```{r}
gg_ls <- list() 

for (sample in unique(spx_meta$sample_id)) {
  
  spx_sample <- spx_meta |>
    dplyr::filter(sample_id == sample)
  nCells_sample <- unique(spx_sample$nCell_sampleSum)
  
  col_max <- signif(max(spx_sample$nCount_RNA) * 2/3, 2)
  g1 <- spx_sample |>  
    dplyr::arrange(nCount_RNA) |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = nCount_RNA)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_gradient(low = "grey75", high = "blue",
                         limits = c(0, col_max), oob = scales::squish) +
    ggtitle(paste0("Cell count (pre-QC): ", nCells_sample)) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed()
  
  col_max <- signif(max(spx_sample$nFeature_RNA) * 2/3, 2)
  g2 <- spx_sample |>  
    dplyr::arrange(nFeature_RNA) |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = nFeature_RNA)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_gradient(low = "grey75", high = "blue",
                         limits = c(0, col_max), oob = scales::squish) +
    theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed()
  
  gg_ls[[sample]] <- (g1 | g2)
  
}
```

```{r, results = 'asis'}
for(i in 1:length(gg_ls)) {
  
  cat("#### ", names(gg_ls)[i], "\n")

  cat("\n")
  print(gg_ls[[i]])
  cat("\n")
  cat("\n")
  
} 
```




## Distribution of total transcripts per cell {.tabset}

Centrality metrics for `nCount_RNA` (before QC) across different groups:

```{r}
spx_meta |>
  group_by(group) |>
  summarise(mean_nCount_RNA = round(mean(nCount_RNA, na.rm =TRUE), 1),
            median_nCount_RNA = round(median(nCount_RNA, na.rm =TRUE), 1),
            q05_nCount_RNA = round(quantile(nCount_RNA, 0.05, na.rm =TRUE), 1),
            q95_nCount_RNA = round(quantile(nCount_RNA, 0.95, na.rm =TRUE), 1))
```

***

**Number of Pass/Fail per sample at user-defined threshold for `nCount_RNA` per cell:**

```{r}
nCount_RNA_min <- min(spx_meta$nCount_RNA[spx_meta$qcFlagCell_summary == "Pass"])
print(paste0("Lowest `nCount_RNA` per cell in cells that passed QC: ", 
             round(nCount_RNA_min, 3)))

# Reminder: by default, the cutoff is set to 20 (unless otherwise specified in the YAML file), as recommended by Bruker for the 1000-plex panel
c20 <- table(spx_meta$sample_id, spx_meta$qcFlagCell_nCount)
c20percent <- round(prop.table(c20, margin = 1) * 100, 1)

df <- cbind(c20, c20percent)
colnames(df) <- c("Total failed cells", "Total passed cells",
                  "Percent failed cells", "Percent passed cells")
df |> as.data.frame()
```

***

In summary, per condition of interest:

```{r}
all_groups <- round(table(spx_meta$qcFlagCell_nCount) / nrow(spx_meta) * 100, 1)
t <- table(spx_meta$group, spx_meta$qcFlagCell_nCount)
t <- round(t / rep(table(spx_meta$group), 2) * 100, 1)
df <- rbind(all_groups, t); df
```


### Per sample

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(nCount_RNA) |>
    ggplot(aes(x = nCount_RNA, y = sample_id, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("nCount_RNA") + 
    geom_vline(xintercept = floor(nCount_RNA_min), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### Per group

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(nCount_RNA) |>
    ggplot(aes(x = nCount_RNA, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("nCount_RNA") + 
    geom_vline(xintercept = floor(nCount_RNA_min), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### Per group (log-scale)

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(nCount_RNA) |>
    ggplot(aes(x = nCount_RNA, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("nCount_RNA") + 
    geom_vline(xintercept = floor(nCount_RNA_min), lty = 2) +
    theme_classic(base_size = font_size) +
    scale_x_log10() +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### `cell_key` rank

```{r, out.width="80%"}
spx_meta_ls <- split(spx_meta, f = spx_meta$slide_id)
for (slide in names(spx_meta_ls)) {
  spx_slide <- spx_meta_ls[[slide]]
  spx_slide <- spx_slide |>
    dplyr::arrange(desc(nCount_RNA)) |> 
    dplyr::mutate(barcode_rank = seq(1:nrow(spx_slide)) / nrow(spx_slide))
  spx_meta_ls[[slide]] <- spx_slide
}
spx_meta <- do.call(rbind.data.frame, spx_meta_ls)

spx_meta |>
  dplyr::arrange(desc(nCount_RNA)) |>
  ggplot(aes(color = slide_id, y = nCount_RNA, x = barcode_rank)) + 
  geom_point() +
  geom_hline(yintercept = floor(nCount_RNA_min), lty = 2, col = "black") +
  scale_y_log10() + 
  theme_minimal() +
  scale_color_manual(values = 
                       colsD) +
  ggtitle("Cells ranked by total transcripts (nCount_RNA)")
```

***


## Distribution of total distinct probes per cell {.tabset}

Centrality metrics for `nFeature_RNA` (before QC) across different groups:

```{r}
spx_meta |>
  group_by(group) |>
  summarise(mean_nFeature_RNA = round(mean(nFeature_RNA, na.rm =TRUE), 1),
            median_nFeature_RNA = round(median(nFeature_RNA, na.rm =TRUE), 1),
            q05_nFeature_RNA = round(quantile(nFeature_RNA, 0.05, na.rm =TRUE), 1),
            q95_nFeature_RNA = round(quantile(nFeature_RNA, 0.95, na.rm =TRUE), 1))
```

***

**Number of Pass/Fail per sample at user-defined threshold for `nFeature_RNA` per cell:**

```{r}
nFeature_RNA_min <- min(spx_meta$nFeature_RNA[spx_meta$qcFlagCell_summary == "Pass"])
print(paste0("Lowest `nFeature_RNA` per cell in cells that passed QC: ", 
             round(nFeature_RNA_min, 3)))

# Reminder: by default, the cutoff is set to 5 (unless otherwise specified in the YAML file), as recommended by Bruker for the 1000-plex panel
c20 <- table(spx_meta$sample_id, spx_meta$qcFlagCell_nFeature)
c20percent <- round(prop.table(c20, margin = 1) * 100, 1)

df <- cbind(c20, c20percent)
colnames(df) <- c("Total failed cells", "Total passed cells",
                  "Percent failed cells", "Percent passed cells")
df |> as.data.frame()
```

***

In summary, per condition of interest:

```{r}
all_groups <- round(table(spx_meta$qcFlagCell_nFeature) / nrow(spx_meta) * 100, 1)
t <- table(spx_meta$group, spx_meta$qcFlagCell_nFeature)
t <- round(t / rep(table(spx_meta$group), 2) * 100, 1)
df <- rbind(all_groups, t); df
```


### Per sample

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(nFeature_RNA) |>
    ggplot(aes(x = nFeature_RNA, y = sample_id, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("nFeature_RNA") + 
    geom_vline(xintercept = floor(nFeature_RNA_min), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### Per group

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(nFeature_RNA) |>
    ggplot(aes(x = nFeature_RNA, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("nFeature_RNA") + 
    geom_vline(xintercept = floor(nFeature_RNA_min), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### Per group (log-scale)

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(nFeature_RNA) |>
    ggplot(aes(x = nFeature_RNA, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("nFeature_RNA") + 
    geom_vline(xintercept = floor(nFeature_RNA_min), lty = 2) +
    theme_classic(base_size = font_size) +
    scale_x_log10() +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### nCount (transcripts) vs. nFeature (probes)

```{r, out.width="80%"}
spx_meta |> 
  ggplot(aes(x=nCount_RNA, y=nFeature_RNA#, color=patient_id
  )) + 
  geom_point(size = 0.5) + 
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = params$cell_nCount_cutoff, col = "red") + 
  geom_hline(yintercept = params$cell_nFeature_cutoff, col = "red") + 
  ggtitle("Probes vs. Transcripts") +
  facet_wrap(~slide_id) +
  xlab("nCount_RNA (log10 scale)")
```

***

## Proportion of negative counts {.tabset}

Centrality metrics for `percentNegCounts` (before QC) across different groups:

```{r}
spx_meta |>
  group_by(group) |>
  summarise(mean_percentNegCounts = round(mean(percentNegCounts, na.rm = TRUE), 1),
            median_percentNegCounts = round(median(percentNegCounts, na.rm = TRUE), 1),
            q05_percentNegCounts = round(quantile(percentNegCounts, 0.05, na.rm = TRUE), 1),
            q95_percentNegCounts = round(quantile(percentNegCounts, 0.95, na.rm = TRUE), 1))
```

***

```{r}
percentNegCounts_max <- max(spx_meta$percentNegCounts[spx_meta$qcFlagCell_summary == "Pass"])
print(paste0("Highest `percentNegCounts` per cell in cells that passed QC: ", 
             round(percentNegCounts_max, 3)))
```

***

Pass/Fail summary, per condition of interest:

```{r}
all_groups <- round(table(spx_meta$qcFlagCell_percentNeg) / nrow(spx_meta) * 100, 1)
t <- table(spx_meta$group, spx_meta$qcFlagCell_percentNeg)
t <- round(t / rep(table(spx_meta$group), 2) * 100, 1)
df <- rbind(all_groups, t); df
```


### Per sample

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(percentNegCounts) |>
    ggplot(aes(x = percentNegCounts, y = sample_id, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("Percentage of negative counts per cell") + 
    geom_vline(xintercept = floor(percentNegCounts_max), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### Per group

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(percentNegCounts) |>
    ggplot(aes(x = percentNegCounts, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("Percentage of negative counts per cell") + 
    geom_vline(xintercept = floor(percentNegCounts_max), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***

### Per group (log-scale)

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(percentNegCounts) |>
    ggplot(aes(x = percentNegCounts, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("Percentage of negative counts per cell") + 
    geom_vline(xintercept = floor(percentNegCounts_max), lty = 2) +
    theme_classic(base_size = font_size) +
    scale_x_log10() +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***


## Cell complexity {.tabset}

The total counts (transcripts) must exceed the number of detected probes (genes) in the cell, i.e. `nCount_RNA` / `nFeature_RNA` > 1

***

Centrality metrics for `complexity` (before QC) across different groups:

```{r}
spx_meta |>
  group_by(group) |>
  summarise(mean_complexity = round(mean(complexity, na.rm = TRUE), 1),
            median_complexity = round(median(complexity, na.rm = TRUE), 1),
            q05_complexity = round(quantile(complexity, 0.05, na.rm = TRUE), 1),
            q95_complexity = round(quantile(complexity, 0.95, na.rm = TRUE), 1))
```

***

```{r}
#spx_meta[is.na(spx_meta$complexity), ]  # one cell with only neg probes
complexity_min <- min(spx_meta$complexity[spx_meta$qcFlagCell_summary == "Pass"], na.rm = TRUE)
print(paste0("Lowest `complexity` per cell in cells that passed QC: ", 
             round(complexity_min, 3)))
```

***

Pass/Fail summary, per condition of interest:

```{r}
all_groups <- round(table(spx_meta$qcFlagCell_complex) / nrow(spx_meta) * 100, 1)
t <- table(spx_meta$group, spx_meta$qcFlagCell_complex)
t <- round(t / rep(table(spx_meta$group), 2) * 100, 1)

df <- rbind(all_groups, t)
df
```


### Per sample

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(complexity) |>
    ggplot(aes(x = complexity, y = sample_id, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("Complexity") + 
    geom_vline(xintercept = floor(complexity_min), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***
***
***

### Per group

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(complexity) |>
    ggplot(aes(x = complexity, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("Complexity") + 
    geom_vline(xintercept = floor(complexity_min), lty = 2) +
    theme_classic(base_size = font_size) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***
***
***

### Per group (log-scale)

```{r}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(complexity) |>
    ggplot(aes(x = complexity, y = group, fill = group)) +
    geom_violin() + geom_boxplot(fill = "white", width = 0.15) + 
    #geom_jitter(width = 0, height = 0.1, size = 0.2) +
    xlab("Complexity") + 
    geom_vline(xintercept = floor(complexity_min), lty = 2) +
    theme_classic(base_size = font_size) +
    scale_x_log10() +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")

  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol))
```

***
***
***


# Conclusion on cell QC

In summary, `spatialhub` flags a cell as passing QC if it fulfills all of the following criteria:

- `nCount_RNA` >= `r params$cell_nCount_cutoff`;
- `nFeature_RNA` >= `r params$cell_nFeature_cutoff`;
- `percentNegCounts` < `r params$cell_percentNeg_cutoff`%;
- `complexity` > `r params$cell_complexity_cutoff`;

Following cell (and if applicable, FOV) QC, **`spatialhub` flags a _sample_ as passing QC if it fulfills both of the following criteria**:

- The number of cells left for that sample after excluding low-quality cells (and if `exclCellsInFailedFOVs` is set to `true`, after excluding low-quality FOVs as well) is >=`r params$sampleMinCells`;
- The percentage of cells left for that sample after excluding low-quality cells (and if `exclCellsInFailedFOVs` is set to `true`, after excluding low-quality FOVs as well) is >`r params$sampleMinPercent`.

In this version of the QC, `exclCellsInFailedFOVs` = `r params$exclCellsInFailedFOVs`.


## Spatial plot flagging Pass/Fail {.tabset}

### low-quality cells highlighted

```{r, fig.height=12L}
spx_meta$qcFlagCell_summary <- factor(as.character(spx_meta$qcFlagCell_summary), levels = c("Pass", "Fail"))

gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(qcFlagCell_summary) |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = qcFlagCell_summary)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_manual(values = alpha(c("grey70", "red2"), 0.5)) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed() + theme(legend.position = 'none')
  
  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol_spatial))
```

***

### low-quality samples flagged

```{r, fig.height=12L}
spx_meta$qcFlagSample_summary <- factor(as.character(spx_meta$qcFlagSample_summary), levels = c("Pass", "Fail"))

gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::arrange(qcFlagSample_summary) |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = qcFlagSample_summary)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_manual(values = alpha(c("grey70", "red2"), 0.5)) +
    ggtitle(slide) + theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed() + theme(legend.position = 'none')
  
  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol_spatial))
```

***

### low-quality cells and samples removed

```{r, fig.height=12L}
gg_slide <- list()
for (slide in sort(unique(spx_meta$slide_id))) {

  g <- spx_meta |>
    dplyr::filter(slide_id == slide) |>  
    dplyr::filter(qcFlagCell_summary == "Pass") |> 
    dplyr::filter(qcFlagSample_summary == "Pass") |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = qcFlagCell_summary)) +
    geom_point(size = 0.02) + theme_classic(base_size = 14) +
    scale_color_manual(values = alpha(c("grey60"), 0.8)) +
    coord_fixed() + theme(legend.position = 'none')
  
  gg_slide[[slide]] <- g
  
}
do.call("grid.arrange", c(gg_slide, ncol = plot_ncol_spatial))
``` 

***


## Summary metrics in filtered dataset {.tabset}

**Per-sample metrics**:

```{r}
qc_vars <- names(spx_meta)
qc_vars <- c("sample_id", "group",
             "nCell_sampleSum", "nCell_sampleSum_passQC", "samplePercentCellsPassQC",
             "samplePercentCellsInFailedFOV", "qcFlagSample_summary",
             qc_vars[grep("sampleSum", qc_vars)], qc_vars[grep("sampleAvg", qc_vars)])
qc_vars <- unique(qc_vars[qc_vars %in% names(spx_meta)])

qc_sample <- spx_meta[, qc_vars]
qc_sample <- qc_sample[!duplicated(qc_sample), ]
rownames(qc_sample) <- 1:nrow(qc_sample)

qc_sample |>
  dplyr::select(all_of(qc_vars)) |>
  dplyr::arrange(desc(nCell_sampleSum_passQC))
```

***

Or, in summary:

```{r}
print("Percentage of cells passing QC:")
spx_meta$qcFlagOverall_summary <- ifelse(spx_meta$qcFlagCell_summary == "Pass" & spx_meta$qcFlagSample_summary == "Pass", "Pass", "Fail")

all_groups <- round(table(spx_meta$qcFlagOverall_summary) / nrow(spx_meta) * 100, 1)
t <- table(spx_meta$group, spx_meta$qcFlagOverall_summary); t <- round(t / rep(table(spx_meta$group), 2) * 100, 1)

df <- rbind(all_groups, t); df
```

***

```{r}
print("Number of cells passing QC:")
all_groups <- table(spx_meta$qcFlagOverall_summary)
t <- table(spx_meta$group, spx_meta$qcFlagOverall_summary)

df <- rbind(all_groups, t); df
```

***

After excluding low-quality cells, the distribution of number of transcripts (`nCount_RNA`) and distinct probes (`nFeature_RNA`) detected per cell looks like this:

```{r}
filtered_meta <- spx_meta |> dplyr::filter(qcFlagOverall_summary == "Pass") 
filtered_meta |>
  group_by(group) |>
  summarise(mean_nCount_RNA = mean(nCount_RNA), mean_nFeature_RNA = mean(nFeature_RNA),
            median_nCount_RNA = median(nCount_RNA), median_nFeature_RNA = median(nFeature_RNA))
```


### density plot - `nCount_RNA`

```{r, out.width="80%", fig.height=6L}
filtered_meta |>
  dplyr::filter(qcFlagOverall_summary == "Pass") |>
  ggplot(aes(x = nCount_RNA, color = slide_id, fill =  slide_id)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = median(filtered_meta$nCount_RNA), lty = 1) +
  geom_vline(xintercept = mean(filtered_meta$nCount_RNA), lty = 2) +
  ggtitle(paste0("Total transcripts per cell")) +
  facet_wrap(. ~ group) +
  NULL
```

***
***
***

### density plot - `nFeature_RNA`

```{r, out.width="80%", fig.height=6L}
filtered_meta |>
  dplyr::filter(qcFlagOverall_summary == "Pass") |>
  ggplot(aes(x = nFeature_RNA, color = slide_id, fill =  slide_id)) + 
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = median(filtered_meta$nFeature_RNA), lty = 1) +
  geom_vline(xintercept = mean(filtered_meta$nFeature_RNA), lty = 2) +
  ggtitle(paste0("Total distinct probes per cell")) +
  facet_wrap(. ~ group) +
  NULL
```

***
***
***



## Total counts and distinct probes per cell (post-QC): per-sample Zoom {.tabset}

```{r}
gg_ls <- list() 

for (sample in unique(filtered_meta$sample_id)) {
  
  spx_sample <- filtered_meta |>
    dplyr::filter(sample_id == sample)
  nCells_sample <- unique(spx_sample$nCell_sampleSum_passQC)
  
  col_max <- signif(max(spx_sample$nCount_RNA) * 2/3, 2)
  g1 <- spx_sample |>  
    dplyr::arrange(nCount_RNA) |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = nCount_RNA)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_gradient(low = "grey75", high = "blue",
                         limits = c(0, col_max), oob = scales::squish) +
    ggtitle(paste0("Cell count (post-QC): ", nCells_sample)) + 
    theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed()
  
  col_max <- signif(max(spx_sample$nFeature_RNA) * 2/3, 2)
  g2 <- spx_sample |>  
    dplyr::arrange(nFeature_RNA) |> 
    ggplot(aes(x = CenterX_composite, y = CenterY_composite, color = nFeature_RNA)) +
    geom_point(size = 0.1) + theme_classic(base_size = font_size) +
    scale_color_gradient(low = "grey75", high = "blue",
                         limits = c(0, col_max), oob = scales::squish) +
    theme(plot.title = element_text(hjust = 0.5)) +
    coord_fixed()
  
  gg_ls[[sample]] <- (g1 | g2)
  
}
```

```{r, results = 'asis'}
for(i in 1:length(gg_ls)) {
  
  cat("### ", names(gg_ls)[i], "\n")

  cat("\n")
  print(gg_ls[[i]])
  cat("\n")
  cat("\n")
  
} 
```


# R session

```{r}
sessionInfo()
```
